Improving Haskell Types with SMT

### 2015-03-13 docs/paper.tex
More paper

\documentclass{sigplanconf}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{fancyvrb}
\usepackage{url}

@@

\conferenceinfo{CONF 'yy}{Month d--d, 20yy, City, ST, Country} 
\copyrightyear{2015} 
\copyrightdata{978-1-nnnn-nnnn-n/yy/mm} 
\doi{nnnnnnn.nnnnnnn}

\title{Title Text}

@@

\begin{abstract}
We present a technique for integrating GHC's type-checker with the
an SMT solver.  The technique was developed to add support for reasoning
about type-level functions on natural numbers, and so our implementation
uses the theory of linear arithmetic.  The technique, however, is more
general and makes it possible to experiment with other external decision
procedures, such as reasoning about type-level booleans, bit-vectors,
or any other theory supported by SMT solvers.
\end{abstract}

@@

XXX: Structure of the paper

@@

\section{Integrating GHC with an SMT Solver}

@@

@@

To check for consistency, we assert all constraints, and ask the SMT
solver if the result is satisfiable.  If this is not the case, then
we report and error, otherwise we proceed with the algorithm.  Note that
since we are working with generalized constraints, if we find an
inconsistency, then we are sure that we've detected a problem.  However,
if the SMT solver says that the constraints are satisfiable, then
we now that this is the case from the point of view of our theory,
however the constraints might still be unsatisfiable if one took a global
view of the problem.  Consider, for example, \Verb"(F a + 4) ~ 7".
This will get imported into our plug-in as \Verb"(x + 4) ~ 7", which is
satisfiable by \Verb"x = 3".  Now, some other theory might know that
\Verb"F a ~ 3" is not actually possible, but we won't detect this problem
at this stage.

@@

\section{Other Theories}

@@

Relation between type-checker and
the Nelson-Oppen method for combining
decision procedures.

\section {Conclusions}

>>>>

\documentclass{sigplanconf}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{fancyvrb}
\usepackage{url}
\usepackage[utf8]{inputenc}

@@

\conferenceinfo{Haskell Symposium 2015}{September, 2015, Vancouver, Canada} 
\copyrightyear{2015} 
\copyrightdata{978-1-nnnn-nnnn-n/yy/mm} 
\doi{nnnnnnn.nnnnnnn}

\title{Improving Haskell Types with SMT}

@@

\begin{abstract}
We present a technique for integrating GHC's type-checker with an SMT solver.
The technique was developed to add support for reasoning
about type-level functions on natural numbers, and so our implementation
uses the theory of linear arithmetic.  However, the approach is not
limited to this theory, and makes it possible to experiment with other
external decision procedures, such as reasoning about type-level booleans,
bit-vectors, or any other theory supported by SMT solvers.
\end{abstract}

@@

The paper is organized as follows:  we start with a brief overview
of SMT solvers form a user's perspective (Section~\ref{smt}); this
section may be skipped by readers familiar with the tools.  Then,
in Section~\ref{GHC}, we introduce the basic concepts of GHC's constraints
solver, which is necessary for putting the rest of the paper in context.
Section~\ref{algorithm} contains the details of the algorithm for
integrating an SMT solver with GHC, and Section~\ref{other} explains
other theories that could be added to GHC using the same technique.
Finally, Section~\ref{modular-typechecker} discusses the possibility
of using the ideas in this paper, and previous work on the
implementation of SMT solver, to engineer a modular constraint solver.

@@

\section{Integrating GHC with an SMT Solver}
\label{algorithm}

@@

\paragraph{Natural Numbers.}
We are interested in working with natural number, however, the SMT
decision procedure works with integers.  To work around this mismatch,
we are careful that whenever we declare an SMT variable corresponding
to a type of kind \Verb"Nat", we also add a formula asserting that
the variable is not negative.

\subsection{Communication with the Solver}

Different SMT solvers support different mechanisms for interacting
with other programs.  The most efficient way to interact with a solver
is to link it with the program, and call into the various methods,
using whatever API is exported by the solver.  Unfortunately, this
approach requires commitment to a specific solver as the low-level
API of the solvers differ.

In our work, we wanted to have the flexibility to experiment with
different solvers, so we opted for a slightly less optimal but
considerably more portable approach.  The multitude of SMT languages
has been recognized as a problem by the SMT community, and the
SMT-LIB standard \cite{smtlib2} was developed to address the issue.
A fair number of solvers have support for the standard, so
to communicate with the solver we developed a Haskell library
\cite{simple-smt}, which can interact with an SMT solver process
by using the SMTLIB language.  The library manages the connection
to the solver, and also provides combinators to create terms
in the common SMT theories.  In the future, we may add support for
direct integration with an SMT solver, and the only impact
should be improved performance.

The code samples in the rest of this Section use the following
API for interaction with the solver:
\begin{Verbatim}
data Solver :: *
data Result = Sat | Unsat

solverAssert :: Solver -> Expr -> IO ()
solverCheck  :: Solver -> IO Result
solverPush   :: Solver -> IO ()
solverPop    :: Solver -> IO ()
\end{Verbatim}
These correspond directly to the commands used in the examples in
Section~\ref{smt}.

@@

To check for consistency, we assert all constraints, and ask the SMT
solver if the result is satisfiable.  If this is not the case, then
we report and error, otherwise we proceed with the algorithm.
\begin{Verbatim}
checkConsistent s cs =
  do mapM_ (solverAssert s) cs
     res <- solverCheck s
     return (res == Sat)
\end{Verbatim}

Note that since we are working with generalized constraints, if we find an
inconsistency, then we are sure that we've detected a problem.  However,
if the SMT solver says that the constraints are satisfiable, then
we now that this is the case from the point of view of our theory,
however the constraints might still be unsatisfiable if one took a global
view of the problem.  Consider, for example, \Verb"(F a + 4) ~ 7".
This will get imported into our plug-in as \Verb"(x + 4) ~ 7", which is
satisfiable by \Verb"x = 3".  Now, some other theory might know that
\Verb"F a ~ 3" is not actually possible, but we won't detect this problem
at this stage.

@@

\section{Other Theories}
\label{other}

@@

In Section~\ref{algorithm} we outlined an algorithm for integrating for
using an SMT solver to solver a certain class of constraints, and then
in Section~\ref{other} we observed that it is possible---and potentially
useful---to solve many other constraints.  This raises an interesting
question: is it possible to structure the entire constraint solver for
a complex programming language, such as Haskell or ML, using a modular
approach, similar to the structure of an SMT solver?

We do not know that answer to this question, but the rest of this Section,
we present a little bit about the structure of SMT solvers, and point out
similarities with GHC's constraint solver, which suggests that this might
be a fruitful area for future research.

\subsection{The Nelson-Oppen Method}

An SMT solver uses a collection of independent decision procedures,
and orchestrates them to compute an answer to a given query.  A well-known
technique for combing decision procedures is due to Nelson and Oppen
\cite{NelsonOppen}.  Their algorithm starts with a collection of
orthogonal theories (i.e., the symbols in the theories do not overlap),
and shows how we may simplify terms in the combined language of the
theories using {\em equality propagation}.  The equality propagation procedure
has a lot of similarities to the algorithm presented in this paper,
and this is entirely by design on our part.

First, constraints that mention symbols from multiple theories are
rewritten so that each constraint mentions only symbols in a single theory.
This is done by naming ``foreign'' sub-terms, in the same way we described
earlier in the paper.

Each collection of constraints is sent to the appropriate decision
procedure to be checked for consistency.  If one of the procedures
detects an inconsistency, then we know that the whole collection of
constraints is inconsistent and we may stop.

If the constraints look consistent, then the decision procedures
are given a chance to ``communicate'' with each other via equality
constraints.  This is analogous to the improvement step in what we
described.  A key observation of Nelson and Oppen is that the only
constraints that need to be communicated are equalities between variables.
The intuition behind this observation is that other equalities would
feature symbols that are ``foreign'' to the other decision procedures,
and they would not know what to do with them.

So, if one decision procedure emits an equality constraint that was not
known by some of the others, we propagate this information, and again
check for consistency and further improvement.

Finally, the Nelson-Oppen technique also allows for a more general
form of communication between procedures---a decision procedure may
indicate that it can solve a problem in multiple ways, by emitting
a {\em disjunction} of equality constraints.  If this happens,
then the coordinating logic has to explore all possibilities, the problem being
satisfiable if any one of the options succeeds, and it is not satisfiable
otherwise.  Searching the options introduced by disjunctions is typically
implemented with the aid of a SAT solver, which is usually considerably
more efficient than simply trying all possibilities, as SAT solvers
propagate information to prune the search space.

To implement the backtracking search efficiently, the decision procedures
are typically incremental (i.e, they can revert back to a previous
known state).  Some techniques can improve this search process further,
by using decision procedures that also produce proofs, \cite{cvc4-lin-arith},
as these proofs can be used to identify dependencies between variables.

So, the overall setups is that we have a collection of independent
decision procedures---each specializing in a single theory---and
they communicate by disjunctions of equalities.  There are three cases,
depending on the shape of the disjunction:
\begin{itemize}
\item an empty disjunction indicates a conflict,
\item a singleton disjunction is a certain improvement, and we propagate
it across theories,
\item a proper disjunction results in search.
\end{itemize}

\subsection{Theories of Haskell}

GHC's constraint solver is not structured exactly in the same way as outlined
in the Nelson-Oppen simplification procedure, there are many similarities.
In particular, equalities play a central role in disseminating information,
but GHC currently does not do any back-tracking search (i.e., decision
procedure may only report inconsistency, or certain knowledge).  However,
there are situations where backtracking search is potentially useful.
For example, when solving class constraints in the context of overlapping
instances, or instance chains \cite{instance-chains}, it is useful
to consider the context of instances, but back-track and try a different
instance, if the context is found to be unsatisfiable.

GHC's implementation of Haskell contains numerous extensions, and so
there are many candidates for potential ``theories'', that is, constraints
that are solved by a dedicated decision procedure.  Engineering a
constraint solver in this style would be very beneficial, both
as one could reason about the correctness of each decision separately,
but also, because it would make it simpler to experiment with new
extensions to the system.

\section {Conclusion}

The paper introduces the basics of GHC's constraint solver, and shows
how to connect it with an external decision procedure that produces
satisfying assignments.  The initial motiviation for this work was
to improve the type-level support for natural numbers, but the solution
turned out to be more general, and opens up the door for intersting
new extentions to Haskell's type system.  More generally, we have
identified similarities between the techniques used to implement SMT solvers,
and the implementaion of GHC's constraint solver, which suggests
the possibility of a modular constraint solver design.
